---
title: "ABS Commonwealth Park flying-fox mapping"
author: "Belinda Wilson and Shoshana Rapley"
date: "11 December 2022"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
editor_options:
  chunk_output_type: console
  
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'tutorial.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

# **Setup**

First, we installed the [pacman Package Management Tool](https://cran.r-project.org/web/packages/pacman/index.html), which allows us to install and load subsequent packages in a condensed and efficient way.

```{r, eval=FALSE}
#install.packages("pacman")
```

```{r, results='hide', warning=FALSE, message=FALSE}
# Install and load required packages
pacman::p_load(plyr, data.table, dplyr, ggmap, ggplot2, janitor, 
               lubridate, naniar, readxl, reshape2, tidyr, tidyverse)
```

# **Species and demographics**

Line graph for 2022 with month on the x-axis and and two y-axes (see below), showing the following as separate lines (Fig 3a):

  1. Number of GHFFs over time (primary axis ranging 0-10,000)
  2. Number of LRFFs over time (primary axis ranging 0-10,000)
  3. Number of trees occupied (secondary axis ranging 0-160)

```{r, results='hide', warning=FALSE, message=FALSE}
# Assign raw data filename to an object
raw_data <- "data.xlsx"

data <- read_excel(raw_data, sheet="monitoring") |>
  clean_names() |>
  filter(include=="Yes", location=="Commonwealth Park")|>
  mutate(lrff=as.character(lrff)) |>
  pivot_longer(cols=9:10, names_to="species", values_to="count") |>
  filter(count!=0) |>
  mutate(count=if_else(count=="800-1000", "900", count)) |>
  mutate(count=if_else(count=="200-250", "225", count)) |> #replaces ranges with mean
  mutate(count=as.numeric(count),
         trees_occupied=as.numeric(trees_occupied),
         date=as.Date(date, format=c("%d/%m/%Y"))) |>
  mutate(year=year(date))

data_22 <- subset(data, date > "2021-12-01" & date < "2022-11-30")

fig3a <- ggplot(data_22) +
  geom_line(mapping=aes(date, count, col=species), size=0.6) +
  geom_line(mapping=aes(date, trees_occupied*50, 
                col='chartreuse4'), size=0.6) +
  scale_color_manual(name='', 
                     values=c("#CACBE8","#1F2148","#7B7FC7"), 
                     labels=c("Trees occupied", 
                              "Grey-headed flying-fox", 
                              "Little red flying-fox")) +
  scale_x_date(date_breaks="1 month", date_labels="%b") +
  scale_y_continuous(breaks=seq(0, 10000, 2000), 
                     sec.axis=sec_axis(~ ./50, 
                      name="Trees occupied")) +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        legend.text=element_text(size=8),
        legend.position="bottom") +
  labs(x="Month", y="Flying-foxes recorded")
print(fig3a)
```

```{r, include=FALSE}
jpeg(file="figures/Flying-fox species and demographics 2022 (Figure 3a).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(fig3a)
dev.off()
```

# **Mean abundance**

Bar graph for 2022 with month on the x-axis and number of flying-foxes on the y-axis, showing the following as separate bars with standard error bars (Fig 3b): 

  1. Mean number of flying-foxes recorded (primary axis ranging 0-6000)
  2. Mean trees occupied (secondary axis ranging 0-140)
  3. Mean ratio of flying-foxes to trees (secondary axis ranging 0-140)

```{r}
data22 <- read_excel(raw_data, sheet="monitoring") |> 
  clean_names() |> 
  filter(include=="Yes", location=="Commonwealth Park", 
         survey!="Estimate") |> #remove estimates 
  mutate(year=year(date)) |> 
  subset(date > "2021-12-01" & date < "2022-11-30") |> 
  mutate(month=lubridate::month(date, label=TRUE, abbr=TRUE), 
         ghff=as.numeric(ghff), 
         lrff=as.numeric(lrff), 
         bats_total=ghff + lrff, 
         ratio=as.numeric(bats_total/trees_occupied)) |> 
  mutate(ratio=ifelse(ratio=="NaN", 0, ratio)) 

#for the purposes of plotting, to help get mean trees 
#and ratio on a secondary axis, we multipled these 
#values by 40 - we undo this at the plotting stage

data22_means <- data22 |> 
  group_by(month) |> 
  summarise(count_mean=mean(bats_total),
            ratio_mean=40*mean(ratio),
            trees_mean=40*mean(trees_occupied)) |>
  pivot_longer(2:4, names_to="type", values_to="mean")

data22_se <- data22 |> 
  group_by(month) |> 
  summarise(count_se=sd(bats_total)/sqrt(length(bats_total)),
            ratio_se=40*sd(ratio)/sqrt(length(ratio)),
            trees_se=40*sd(trees_occupied)/sqrt(length(trees_occupied))) |>
  pivot_longer(2:4, names_to="type", values_to="se")

data22_stats <- cbind(data22_means, data22_se)
colnames(data22_stats) <- make.unique(names(data22_stats))

june <- data.frame("Jun", "count_mean", NA, "Jun", "count_se", NA)
july <- data.frame("Jul", "count_mean", NA, "Jul", "count_se", NA)
august <- data.frame("Aug", "count_mean", NA, "Aug", "count_se", NA)
september <- data.frame("Sep", "count_mean", NA, "Sep", "count_se", NA)

names(june) <- c("month", "type", "mean", "month.1", "type.1", "se")
names(july) <- c("month", "type", "mean", "month.1", "type.1", "se")
names(august) <- c("month", "type", "mean", "month.1", "type.1", "se")
names(september) <- c("month", "type", "mean", "month.1", "type.1", "se")

data22_stats <- rbind(data22_stats, june, july, august, september) |>
  mutate(type=factor(type,levels=c("count_mean", "trees_mean", 
                                   "ratio_mean", "count_se", 
                                   "trees_se", "ratio_se")),
         month=factor(month, levels=c("Dec", "Jan", "Feb", "Mar", 
                                      "Apr", "May", "Jun", "Jul", 
                                      "Aug", "Sep", "Oct", "Nov")))

#data22_stats$true <- data22_stats$mean/40 #true values for report

fig3b <- ggplot(data22_stats, aes(x=month, y=mean, fill=type)) +
  geom_bar(stat="identity", position="dodge") + 
  geom_errorbar(mapping=aes(x=month, ymin=mean-se, 
                            ymax=mean+se, col=type), 
                position="dodge") + 
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        legend.text=element_text(size=8),
        legend.position="bottom") +
  scale_fill_manual(name='', values=c("#1F2148","#CACBE8","#7B7FC7"),
                    labels=c("Mean flying-fox abundance", 
                             "Mean trees occupied", 
                             "Mean ratio of flying-fox to trees")) +
  scale_y_continuous(sec.axis=sec_axis(~ ./40, 
                    name="Mean trees occupied and \nmean ratio of flying-fox to trees")) +
  scale_colour_manual(name='', guide="none",
                    values=c("#CACBE8","#7B7FC7","#1F2148")) +
  labs(x="Month", y="Mean flying-fox abundance")
print(fig3b)
```

```{r, include=FALSE}
jpeg(file="figures/Mean flying-fox abundance 2022 (Figure 3b).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(fig3b)
dev.off()
```

# **Peak numbers**

Stacked bar graph for 2012-2022 with year on the x-axis and number of flying-foxes recorded on the y-axis (Fig 4), showing:

  1. Maximum number of flying-foxes recorded for each year
  2. Mean number of flying-foxes recorded for each year

```{r}
data <- read_excel(raw_data, sheet="monitoring") |>
  clean_names() |>
  drop_na(ghff)|>
  filter(include=="Yes", location=="Commonwealth Park") |>
  mutate(ghff=if_else(ghff=="800-1000", "900", ghff)) |>
  mutate(ghff=if_else(ghff=="200-250", "225", ghff)) |> #replaces ranges with mean
  mutate(ghff=as.numeric(ghff),
         lrff=as.numeric(lrff),
         year=year(date)) |> 
  mutate(bats_total=ghff + lrff)

peak <- data |> 
  group_by(year) |>
  summarise(bats_mean=mean(bats_total), 
            bats_max=max(bats_total)) |>
  mutate(bats_diff=bats_max-bats_mean) |>
  pivot_longer(2:4, names_to="type", values_to="stat")
peak <- filter(peak, type!="bats_max")

fig4 <- ggplot(peak, (aes(year, stat, fill=type))) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        legend.text=element_text(size=8),
        legend.position="bottom") +
  scale_fill_manual(name="", values=c("#1F2148","#CACBE8"),
                    labels=c("Peak", "Mean")) +
  scale_y_continuous(breaks=seq(0, 10000, 2000)) +
  scale_x_continuous(breaks=seq(2011, 2022, 1)) +
  labs(x="Year", y="Flying-fox abundance") +
  theme(plot.title=element_text(size=15, face="bold", 
                                margin=margin(10, 0, 10, 0)))
print(fig4)
```

```{r, include=FALSE}
jpeg(file="figures/Peak flying-fox abundance 2012–2022 (Figure 4).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(fig4)
dev.off()
```

# **Survey effort and weather**

Line graph for 2012-2022 with date on the x-axis and number of flying-foxes recorded on the y-axis, showing the number of flying-foxes recorded for every survey conducted in each year, with temperature in the background (Fig 5).

The weather data were acquired from the [Australian Bureau of Meteorology weather station 70351](http://www.bom.gov.au/climate/data/index.shtml?bookmark=136).

```{r}
weather <- read_excel(raw_data, sheet="weather") |>
  clean_names() |> 
  filter(year>=2011) |>
  mutate(date=paste(year, month, day, sep="/")) |>
  mutate(date=as.Date(date, format=c("%Y/%m/%d")))

weather2 <-pivot_longer(weather, 5:6, 
                        names_to="type", 
                        values_to="temperature")

fig5 <- ggplot() +
  geom_ribbon(data=weather, aes(as.Date(date), 
                                ymin=temp_min, ymax=temp_max), 
              fill="#CACBE8", alpha=0.5) +
  #geom_point(data=weather2, aes(date, temperature, colour=temperature)) +
  #scale_colour_gradient2(low="darkblue", mid="white", high="coral4",
                         #aesthetics="colour", midpoint=15) +
  geom_line(data=data, aes(as.Date(date), bats_total/200), 
            size=0.6, col="#1F2148") +
  scale_y_continuous(name="Daily temperature (°C)", 
                     sec.axis=sec_axis(~ .*200, 
                                       breaks=seq(0, 10000, 2000),
                                       name="Flying-fox abundance")) +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank()) +
  labs(x="Year", y="Temperature")
print(fig5)
```

```{r, include=FALSE}
jpeg(file="figures/Survey effort and weather 2012–2022 (Figure 5).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(fig5)
dev.off()
```

# **Density map**

Here we map the density of flying-foxes in each tree across the camp (Fig 6). `surveys_wide` is from the `surveys` table in the database, and `trees` is from the `trees` table. 

Because we didn't do any surveys in June and July (but want them to appear in our figure) *you may need to add dummy surveys for these months* into the initial `surveys` dataset. You can do this by copying the row for 22/08/2020 (where all the trees have 0 bats) several times for each missing month per year, and changing the date to 1/6/2020, for example. 

  1. Convert tree columns from wide format (one per column) to long format (tree identities listed in one column).

```{r}
surveys_wide <- read_excel("data.xlsx", sheet="surveys") |>
  subset(SurveyID!="N/A")

surveys <- melt(surveys_wide, measure.vars=c(30:203), 
                variable.name="tree", value.name="bats")
```

  2. Merge surveys with coordinates for each tree
  3. Create `month` and `year` variables

```{r, results='hide', warning=FALSE, message=FALSE}
trees <- read_excel(raw_data, sheet="trees") |>
  clean_names()

surveys_trees <- left_join(surveys, trees, by="tree") |>
  clean_names() |>
  subset(bats!=999) |>
  mutate(year=format(date, "%Y"),
         month=format(date, "%b"), 
         dates=format(date, "%d %b %y"))
```

 4. Get base layer from Google Maps

You'll need to sign up for a Google API for this step.

```{r, results='hide', warning=FALSE, message=FALSE}
# Use a Google API to fetch a base map
map <- get_map(location=c(lon=149.131749, lat=-35.289492), 
               zoom=19, source="google", 
               maptype="roadmap", crop=FALSE)
```

  5. Order months
  6. Categorise number of bats (didn't end up using this)
  7. Subset the data for 2019-2022

```{r}
data <- surveys_trees |>
  mutate(date=as.Date(date, format=c("%d/%m/%Y"))) |>
  subset(date > "2021-12-01" & date < "2022-11-30") |>
  mutate(month=factor(month, levels=c("Dec", "Jan", "Feb", 
                                      "Mar", "Apr", "May", 
                                      "Jun", "Jul", "Aug", 
                                      "Sep", "Oct", "Nov")),
         foxes=cut(bats, breaks=c(-Inf, 50, 100, 200, 400, Inf), 
                   labels=c("0-50", "50-100", "100-200", 
                            "200-400", ">400"))) |>
  mutate(foxes=factor(foxes, levels=c(">400", "100-200", 
                                      "200-400", "50-100", 
                                      "0-50")))
```

  6. Map the number of `Bats` recorded per `Month`

```{r}
fig6 <- ggmap(map) + 
  geom_point(data, mapping=aes(x=longitude, y=latitude), 
             shape=8, col="darkgreen", size=1) +
  geom_point(data, size=1, 
             mapping=aes(x=longitude, y=latitude, 
                         col=bats, alpha=bats)) +
  scale_colour_gradient2(low=0, midpoint=10, 
                         mid="yellow", high="red") +
  facet_wrap(~month, ncol=4) +
  theme(panel.grid.major=element_blank(),
        panel.background=element_rect(fill="white"),
        plot.margin=unit(c(0, 0, 0, 0), "cm"),
        axis.text.x=element_blank(), 
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        legend.key=element_blank(),
        strip.background=element_rect(fill="#1F2148"),
        strip.text=element_text(colour="white")) +
  guides(col=guide_legend(title="Number of \nflying-foxes\nrecorded",
                          reverse=TRUE), 
         alpha=guide_legend(title="Number of \nflying-foxes\nrecorded",
                            reverse=TRUE)) +
  xlab("") + ylab("")
print(fig6)
```

```{r, include=FALSE}
jpeg(file="figures/Flying-fox density 2022 (Figure 6).jpeg", 
     width=5000, height=3000, units="px", res=800)
print(fig6)
dev.off()
```

# **Session information**

```{r}
# Display version information for R, OS, and packages
sessionInfo()
```