---
title: "Hanging in there: the highs and lows of a flying-fox camp in a novel landscape"
author: "Wilson B A & Rapley S"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
    top_depth: 3
    toc_float:
      collapsed: true
    theme: cerulean
    highlight: pygments
    self_contained: true
editor_options:
  chunk_output_type: console
knit: "(function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'tutorial.html')) })"
---

# **Background**

This [R project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) presents the datasets and workflows used in the '*Hanging in there: the highs and lows of a flying-fox camp in a novel landscape*' project. For information, visit the [GitHub repo](https://github.com/ecologibel/commonwealth-park-flying-fox-monitoring). 

# **Setup**

## Set global options

```{r global-options, echo = FALSE}
# Set default knitting options 
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = TRUE
)

# Set chunk behaviours (e.g., chunks with 'draft = TRUE' will not run)
knit <- list(
  # For one-off installation chunks (do not run)
  init_once = list(eval = FALSE),
  # For setup chunks (hide chunk and output)
  setup     = list(echo = FALSE, message = FALSE, warning = FALSE), 
  # For setup function chunks (show chunk but not output)
  echo_only = list(echo = TRUE, message = FALSE, warning = FALSE),
  # For function chunks (hide chunk but show output)
  fun       = list(echo = FALSE), 
  # For draft chunks (do not run)
  draft     = list(eval = FALSE),
  # For dynamic summary chunks (hide chunk but show inline results)
  ds        = list(echo = FALSE, results = 'asis'),
  # For plot chunks Plotting chunk (always run)
  plot      = list(eval = TRUE),
  # Set size defaults for printed figures
  fig       = list(fig.height = 10, fig.width = 15),
  # For ggsave chunks (always run) 
  ggsave    = list(eval = TRUE)
)

# Define hook functions based on knit list
knitr::opts_hooks$set(
  draft = function(options) {
    if (isTRUE(options$draft)) {
      options$eval <- FALSE
    }
    options
  }
)
```

## Initialise project

To initialise your project, load the function below then run `run_setup()`. 

Depending on how many packages need installing, it may take a while. Feel free to have a look through `setup.Rmd` so you're familiar with the setup. 

```{r, setup = TRUE}
# Define setup function
run_setup <- function(path = "setup.Rmd", total = 3) {
  if (!requireNamespace("cli", quietly = TRUE)) {
    install.packages("cli")}
  library(cli)

  # Progress bar
  cli_progress_bar("Initialising setup", total = total)

  if (is.null(knitr::opts_knit$get("output.dir"))) {
    knitr::opts_knit$set(output.dir = getwd())}
  cli_progress_update(set = 2)

  if (isTRUE(getOption("knitr.in.progress"))) {
    # During knitting → include setup.Rmd inline 
    # (respects chunk options + text)
    invisible(knitr::knit_child(
      path, 
      quiet = TRUE, 
      envir = knitr::knit_global()))
  } else {
    
    # Interactive use → inline helper to extract & eval code blocks
    extract_r_code <- function(file) {
      lines  <- readLines(file, warn = FALSE)
      starts <- grep("^```\\{r", lines)
      ends   <- grep("^```\\s*$", lines)
      blocks <- mapply(
        function(s, e) {
          header <- lines[s]
          code <- lines[(s + 1):(e - 1)]
          # Skip if eval = FALSE
          if (grepl("eval\\s* = \\s*FALSE", header)) return(NULL)
          code
        },
        starts, 
        ends,
        SIMPLIFY = FALSE
      )
      paste(unlist(blocks), collapse = "\n")
    }

    code <- extract_r_code(path)
    eval(parse(text = code), envir = globalenv())
  }

  cli_progress_update(set = 3)
  cli_progress_done()
  cli_alert_success("Setup successful!")
}
```

```{r}
# Run setup 
run_setup()
```

## Load project-specific packages

This script installs and loads *project-specific* packages (*commonly used* packages are loaded in `setup.Rmd`). 

```{r, results = 'hide', warning = FALSE, message = FALSE}
# Install and load project-related packages
pacman::p_load(
  # Data wrangling
  grid, hms, naniar, reshape2 
) 
```

## Define resources

This chunk tabulates the abbreviations used in our workflows.  

```{r, draft = TRUE}
# Read in abbrevations list and print it as a clean table
(def <- print_clean_table(raw_data, sheet = "definitions"))
```

### Constants

```{r}
# Define dates for current reporting period
START            <- "2024-12-01"
END              <- "2025-11-30"
REPORTING_PERIOD <- 2025

# Define whether LRFFs were recorded in the last reporting period
LRFFS_RECORDED   <- FALSE

# Define visual scaling factor to align trees occupied with FF abundance
TREE_SCALE       <- 20

# Define visual scaling factor to align FF abundance with temperature
FF_SCALE         <- 200
```

```{r}
# Define constant ggplot arguments
map_height  <- 6
map_width   <- 6
plot_height <- 6 
plot_width  <- 9
dpi         <- 300
```

```{r}
# Define columns in `monitoring` sheet
monitoring_cols <- c(
  "location", 
  "survey", 
  "date", 
  "start", 
  "finish", 
  "hours", 
  "ghff", 
  "lrff", 
  "trees_occupied", 
  "mothers_and_pups", 
  "heat_stress", 
  "wind", 
  "cloud", 
  "humidity", 
  "bat_noise", 
  "other_noise", 
  "rubbish", 
  "humans", 
  "include"
)

# Define categorical variables
cat_vars <- c(
  "pups", 
  "heat", 
  "bat_noise", 
  "other_noise", 
  "humans", 
  "rubbish", 
  "cloud", 
  "humid",
  "wind"
)

# Define long labels for categorical variables
cat_labs <- c(
  "pups"        = "Presence of mothers and pups", 
  "heat"        = "Heat stress behaviours", 
  "bat_noise"   = "Flying-fox noise", 
  "other_noise" = "Other noise", 
  "humans"      = "Presence of humans", 
  "rubbish"     = "Presence of litter", 
  "cloud"       = "Cloud cover", 
  "humid"       = "In situ humidity", 
  "wind"        = "Wind strength"
)
```

### Functions

```{r}
# Define function to convert Excel times to hms
to_hms <- function(x) {
  # Leave hms values unchanged
  if (inherits(x, "hms")) {
    return(x)
  }

  # Extract time-of-day from POSIXct/POSIXlt datetimes
  if (inherits(x, "POSIXt")) {
    return(hms::as_hms(as.numeric(x) %% 86400))
  }

  # Convert difftime durations to seconds
  if (inherits(x, "difftime")) {
    return(hms::as_hms(as.numeric(x, units = "secs")))
  }

  # Handle numeric Excel times by keeping the fractional day
  if (is.numeric(x)) {
    frac_day <- x %% 1
    return(hms::as_hms(frac_day * 86400))
  }

  # Parse character times and treat blanks as missing
  x_chr <- na_if(trimws(as.character(x)), "")
  hms::as_hms(readr::parse_time(x_chr, format = "%H:%M"))
}
```

# **Data preparation**

We entered the FF survey data in a [Microsoft Access](https://www.microsoft.com/en-au/microsoft-365/access) database. To put this into a form that we can use in this R markdown: 

  1. Copy the `Survey` and `Tree` tables from the database into their respective sheets in `input/raw data.xlsx`. 
  2. Copy the rows from the last reporting period (Dec 1–Nov 30) in the `Schedule` sheet of the `ABS flying-fox monitoring` Google spreadsheet (requires specific access) to the bottom of the `monitoring` sheet.

```{r}
# Read in data
dat <- read_excel(raw_data, sheet = "monitoring") %>% 
  # Standardise column names to lowercase with underscores
  clean_names() %>%
  # Select relevant columns
  select(all_of(monitoring_cols)) %>%
  # Convert columns
  mutate(
    # Convert dates and times
    across(c(start, finish, hours), to_hms), 
    ghff = as.character(ghff), 
    lrff = as.character(lrff)
  ) %>%
  
  # Filter to relevant site
  filter(
    include  == "Yes", 
    location == "Commonwealth Park"
  ) 
```


# **Plots**

## Abundance and trees occupied (Figure 2a)

First, we generate a **(a)** raw line graph and a **(b)** smoothed line graph by month, with two y-axes:

  1. **Primary axis:** GHFF and LRFF abundance over time (0–10,000)
      - *Note: no LRFFs were recorded during the 2025 reporting period.*
  2. **Secondary axis:** Number of trees occupied by FFs (0–160)

```{r, results = 'hide', warning = FALSE, message = FALSE}
# Pivot df to long format
dat_wide <- dat %>%
  
  pivot_longer(
    cols      = c("ghff", "lrff"), 
    names_to  = "species", 
    values_to = "count"
  ) %>%
  
  # Convert count values
  mutate(count = if_else(count == "800-1000", "900", count)) %>%
  mutate(count = if_else(count == "200-250",  "225", count)) %>% 
  
  # Replace ranges with mean
  mutate(
    count          = as.numeric(count),
    trees_occupied = as.numeric(trees_occupied),
    date           = as.Date(date, format = c("%d/%m/%Y"))
  ) %>%
  mutate(year = year(date))

# Subset to current reporting period
dat_cur <- dat_wide %>%
  filter(date > START & date < END) %>%
  # Include LRFFs if LRFFS_RECORDED = TRUE 
  filter(LRFFS_RECORDED | species != "lrff")

# Export df to a new sheet in the running workbook
workbook <- export_sheet(dat_cur, sheet = "data")
```

### Raw line graph

```{r}
# Line plot raw abundance
(abun_line <- ggplot(dat_cur) +
  
  geom_ribbon(
    aes(
      x    = date,
      ymin = 0,
      ymax = trees_occupied * TREE_SCALE,
      fill = "Trees occupied"
    ),
    alpha  = 0.25,
    colour = NA,
    na.rm  = TRUE
  ) +
   
  geom_line(
    mapping   = aes(
      col     = "Trees occupied", 
      x       = date, 
      y       = trees_occupied * TREE_SCALE
    ), 
    linewidth = 0.7
  ) +
   
  geom_ribbon(
    aes(
      x    = date,
      ymin = 0,
      ymax = count,
      fill = "Flying-foxes recorded"
    ),
    alpha  = 0.25,
    colour = NA
  ) +
   
  geom_line(
    mapping   = aes(
      col     = "Flying-foxes recorded", 
      x       = date, 
      y       = count
    ), 
    linewidth = 0.7
  ) +
  
  labs(
    tag = "a", 
    x   = "Month", 
    y   = "Flying-foxes recorded"
  ) + 
  
  scale_color_manual(
  name   = "",
  values = c(
    "Trees occupied"        = "#458B00",
    "Flying-foxes recorded" = "#7B7FC7"
  )) + 
 
  scale_fill_manual(
  name   = "",
  values = c(
    "Trees occupied"        = "#458B00",
    "Flying-foxes recorded" = "#7B7FC7"
  )) + 
   
  scale_x_date(
    date_breaks = "1 month", 
    date_labels = "%b"
  ) +
  scale_y_continuous(
    sec.axis = sec_axis(~ ./TREE_SCALE, name = "Trees occupied")
  ) + 
  
  theme_minimal() + 
  theme(
    axis.line         = element_line(colour = "darkgrey"), 
    axis.text.x       = element_text(angle = 90, hjust = 1, vjust = 1),
    axis.text.y       = element_text(vjust = 0.5), 
    axis.title.x      = element_blank(), 
    axis.title.y      = element_text(vjust = 2, margin = margin(r = 1)),
    legend.position   = "bottom", 
    legend.title      = element_blank(), 
    panel.background  = element_rect(fill = "white", colour = NA), 
    panel.grid        = element_blank(),
    plot.margin       = margin(t = 0.2, r = 0.5, b = -0.2, l = 0.1, "cm"), 
    plot.tag          = element_text(face = "bold", size = 12), 
    plot.tag.position = c(0.8, 0.98)
  )
)
```

### Smoothed line graph

```{r}
# Smooth plot abundance
(abun_smooth <- ggplot(dat_cur) + 
   
  geom_smooth(
    mapping   = aes(
      col     = "Flying-foxes recorded", 
      fill    = "Flying-foxes recorded", 
      x       = date, 
      y       = count
    ), 
    linewidth = 0.7
  ) +
  
  geom_smooth(
    mapping   = aes(
      col     = "Trees occupied", 
      fill    = "Trees occupied", 
      x       =  date, 
      y       =  trees_occupied * TREE_SCALE
    ), 
    linewidth = 0.7, 
    na.rm     = TRUE
  ) + 
  
  labs(
    tag = "b", 
    x   = "Month", 
    y   = "Flying-foxes recorded"
  ) + 

  scale_color_manual(
  name   = "",
  values = c(
    "Trees occupied"        = "#458B00",
    "Flying-foxes recorded" = "#7B7FC7"
  )) + 
 
  scale_fill_manual(
  name   = "",
  values = c(
    "Trees occupied"        = "#458B00",
    "Flying-foxes recorded" = "#7B7FC7"
  )) + 
    
  scale_x_date(
    date_breaks = "1 month", 
    date_labels = "%b"
  ) +
    
  scale_y_continuous(
    sec.axis = sec_axis(~ ./TREE_SCALE, name = "Trees occupied")
  ) + 
  
  theme_minimal() +
  theme(
    axis.line         = element_line(colour = "darkgrey"), 
    axis.text.x       = element_text(angle = 90, hjust = 1, vjust = 1),
    axis.text.y       = element_text(vjust = 0.5),
    axis.title.x      = element_blank(), 
    axis.title.y      = element_text(vjust = 2, margin = margin(r = 1)),
    legend.text       = element_text(size  = 8),
    legend.position   = "bottom", 
    legend.title      = element_blank(),
    panel.background  = element_rect(fill = "white", colour = NA), 
    panel.border      = element_blank(),
    panel.grid.major  = element_blank(), 
    panel.grid.minor  = element_blank(), 
    plot.margin       = margin(t = 0.2, r = 0.1, b = -0.2, l = 0.1, "cm"), 
    plot.tag          = element_text(face = "bold", size = 12), 
    plot.tag.position = c(0.8, 0.98)
  )
)
```

## Mean abundance per month (Figure 2b)

Bar graph with month on the x-axis and FF abundance on the y-axis, showing the following as separate bars with standard error bars: 

  1. Mean (±SE) FFs recorded (primary axis ranging 0–6,000)
  2. Mean (±SE) trees occupied (secondary axis ranging 0–140)
  3. Mean (±SE) ratio of FFs-to-trees (secondary axis ranging 0–140)

For the purposes of plotting the mean number of trees occupied and tree-by-bat ratio on a secondary axis, we multiplied these values by 40, then reversed this before plotting.

### Data preparation

```{r}
# Read in the data
dat_cur2 <- dat_cur %>%
  
  mutate(
    month      = lubridate::month(
      date, 
      label    = TRUE, 
      abbr     = TRUE
    ), 
    count      = as.numeric(count), 
    ratio      = as.numeric(count / trees_occupied)
  ) %>% 
  
  mutate(ratio = ifelse(ratio == "NaN", 0, ratio)) 

# Calculate mean statistics
dat_cur_mean <- dat_cur2 %>% 
  group_by(month) %>% 
  
  summarise(
    count_mean = mean(count),
    ratio_mean = TREE_SCALE * mean(ratio),
    trees_mean = TREE_SCALE * mean(trees_occupied)
  ) %>%
  
  pivot_longer(
    2:4, 
    names_to = "type", 
    values_to = "mean"
  )

# Calculate standard error (SE) statistics
dat_cur_se <- dat_cur2 %>% 
  group_by(month) %>% 
  
  summarise(
    count_se = sd(count) / 
      sqrt(length(count)),
    
    ratio_se = TREE_SCALE * sd(ratio) / 
      sqrt(length(ratio)),
    
    trees_se = TREE_SCALE * sd(trees_occupied) / 
      sqrt(length(trees_occupied))) %>%
  
  pivot_longer(
    2:4, 
    names_to  = "type", 
    values_to = "se"
  )

# Combine means and SE
dat_cur_stats <- cbind(
  dat_cur_mean, 
  dat_cur_se
)

colnames(dat_cur_stats) <- make.unique(names(dat_cur_stats))
```

For each month, regardless of whether FFs were detected, manually generate rows for these months.

```{r}
# Manually create rows for months where bats were not detected
jul <- data.frame(
  "Jul", 
  "count_mean", 
  NA, 
  "Jul", 
  "count_se", 
  NA
)

names(jul) <- c(
  "month", 
  "type", 
  "mean", 
  "month.1", 
  "type.1", 
  "se"
)

aug <- data.frame(
  "Aug", 
  "count_mean", 
  NA, 
  "Aug", 
  "count_se", 
  NA
)

names(aug) <- c(
  "month", 
  "type", 
  "mean", 
  "month.1", 
  "type.1", 
  "se"
)

sep <- data.frame(
  "Sep", 
  "count_mean", 
  NA, 
  "Sep", 
  "count_se", 
  NA)

names(sep) <- c(
  "month", 
  "type", 
  "mean", 
  "month.1", 
  "type.1", 
  "se"
)

# Combine rows with df
dat_current_stats <- rbind(
  dat_cur_stats, 
  jul, 
  aug, 
  sep
  ) %>%
  
  mutate(
    type = factor(type, levels = c(
      "count_mean", 
      "trees_mean", 
      "ratio_mean", 
      "count_se", 
      "trees_se", 
      "ratio_se"
    )),
    
    month = factor(month, levels = c(
      "Dec", "Jan", "Feb", 
      "Mar", "Apr", "May", 
      "Jun", "Jul", "Aug", 
      "Sep", "Oct", "Nov"
    )
  )
)

# Reverse × 40 multiplier
dat_cur_stats$true <- dat_cur_stats$mean/TREE_SCALE 

# Create a three-level grouping variable for metrics
dat_cur_stats <- dat_cur_stats %>%
  mutate(metric = sub("_(mean|se)$", "", type))
```

### Plot

```{r}
# Generate barplot
(abun_mean <- ggplot(
  
  dat_cur_stats, 
  aes(
    fill = metric, 
    x    = month, 
    y    = mean
  )) +
  
  geom_col(position = "dodge") + 
  
  geom_errorbar(
    mapping = aes(
      col   = metric, 
      x     = month, 
      ymax  = mean + se, 
      ymin  = mean - se
    ), 
    position = "dodge"
  ) + 
  
  labs(
    tag = "c", 
    x   = "Month", 
    y   = "Mean flying-fox abundance"
  ) + 
 
  scale_colour_manual(
  name    = "", 
  breaks  = c("count", "trees", "ratio"), 
  guide   = "none", 
  labels  = c(
    "Mean flying-fox abundance",
    "Mean trees occupied",
    "Mean ratio of flying-fox to trees"),
  values  = c(
    count = "#51549E", 
    trees = "#A1CC76", 
    ratio = "#C4A487")
  ) + 
   
  scale_fill_manual(
  name   = "", 
  breaks = c("count", "trees", "ratio"),
  labels = c(
    "Mean flying-fox abundance",
    "Mean trees occupied",
    "Mean ratio of flying-fox to trees"),
  values  = c(
    count = "#1F2148",
    trees = "#458B00", 
    ratio = "#8B6B4F")
  ) + 
   
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ ./TREE_SCALE, 
      name = "Mean trees occupied and \nmean ratio of flying-fox to trees")
  ) + 
   
  theme_minimal() + 
  
  theme(
    axis.line         = element_line(colour = "darkgrey"), 
    axis.text.x       = element_text(vjust = 1),
    axis.text.y       = element_text(vjust = 0.5),
    axis.title.x      = element_blank(), 
    axis.title.y      = element_text(vjust = 3, margin = margin(r = 1)), 
    legend.position   = "bottom", 
    legend.text       = element_text(size = 8),
    legend.title      = element_blank(),
    panel.background  = element_rect(colour = NA, fill = "white"), 
    panel.grid.major  = element_blank(), 
    panel.grid.minor  = element_blank(), 
    panel.border      = element_blank(),
    plot.margin       = margin(t = 0.5, r = 0.1, b = -0.2, l = 0.3, "cm"), 
    plot.tag          = element_text(face = "bold", size = 12), 
    plot.tag.position = c(0.88, 0.98)
  )
)
```

### Combine plots

```{r, eval = FALSE, include = FALSE}
# Combine plots into a single figure
(abun_raw <- ggarrange(
  abun_line, 
  abun_smooth, 
  ncol    = 2, 
  nrow    = 1, 
  widths  = c(1, 1), 
  align   = "h"
))

# Combine plots into a single figure
(abun_plots <- ggarrange(
  abun_raw, 
  abun_mean, 
  ncol     = 1, 
  heights  = c(1, 1)
))

# Export plot as a jpeg file
ggsave(
  filename = "output/fig 2 current abundance.jpeg",
  plot     = abun_plots, 
  height   = 140, 
  width    = 200, 
  units    = "mm", 
  dpi      = 600
)
```

## Mean abundance per year (Figure 3)

Annual distribution of flying-fox abundance between 2012–current, showing:

  1. Peak (maximum) FFs recorded for each year
  2. Mean FFs recorded for each year

### Data preparation

```{r}
# Prepare the data
dat_all <- dat %>%
  drop_na(ghff) %>%
  
  mutate(ghff = if_else(ghff == "800-1000", "900", ghff)) %>%
  
  # Replace ranges with mean
  mutate(ghff = if_else(ghff == "200-250",  "225", ghff)
  ) %>% 
  
  mutate(
    ghff = as.numeric(ghff),
    lrff = as.numeric(lrff),
    year = year(date)
  ) %>% 
  
  mutate(bats_total = ghff + lrff) %>%
  
  # Remove years prior to year 2011
  filter(year > 2011) %>%
  
  # Calculate peak abundance
  group_by(year) %>%
  mutate(peak = max(bats_total, na.rm = TRUE)) %>%
  ungroup()
```

```{r}
# Bar plot
(abun_year <- ggplot(dat_all, (aes(x = factor(year), y = bats_total))) + 
    
  # Draw violin plot 
  geom_violin(
    aes(col  = peak, 
        fill = peak), 
    alpha    = 0.9, 
    lwd      = 0.6, 
    na.rm    = TRUE, 
    scale    = "width", 
    width    = 0.8
  ) + 

  # Draw boxplot 
  geom_boxplot(
    col           = "white", 
    fill          = "white", 
    lwd           = 0.5, 
    na.rm         = TRUE, 
    outlier.shape = NA, 
    width         = 0.05
  ) + 

  # Remove white ticks on legend scale
  guides(fill = guide_colorbar(ticks.colour = NA)) + 
    
  labs(
    x = "Year", 
    y = "Flying-fox abundance"
  ) + 
  
  scale_colour_gradientn(
    colours        = c("#CACBE8", "#7B7FC7", "#000333"), 
    guide          = guide_colorbar(
      barheight    = unit(3, "cm"), 
      barwidth     = unit(0.5, "cm")
    ), 
    name           = "Peak"
  ) + 
    
  scale_fill_gradientn(
    colours        = c("#CACBE8", "#7B7FC7", "#000333"), 
    guide          = guide_colorbar(
      barheight    = unit(3, "cm"), 
      barwidth     = unit(0.5, "cm")
    ), 
    name           = "Peak"
  ) + 
  
  scale_x_discrete(
    drop = FALSE, 
    breaks = seq(2011, REPORTING_PERIOD, 1)
  ) + 
    
  scale_y_continuous(breaks = seq(0, 10000, 2000)) + 
  
  # Define median value point for boxplot
  stat_summary(
    col    = mono4, 
    fill   = white, 
    fun    = "median", 
    geom   = "point", 
    shape  = 21, 
    size   = 1.7, 
    stroke = 1
  ) + 
    
   # Apply visual theme
  theme_minimal() + 
  theme(
    axis.line        = element_line(colour = "darkgrey"), 
    axis.text.x      = element_text(vjust  = 0.5),
    axis.title.x     = element_blank(), 
    axis.title.y     = element_text(margin = margin(r = 5), vjust = 0.5),
    legend.position  = c(0.07, 0.7), 
    legend.text      = element_text(size = 8), 
    legend.title     = element_text(face = "bold", size = 8), 
    panel.background = element_rect(fill = "white", colour = NA), 
    panel.border     = element_blank(),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    plot.margin      = margin(t = -0.2, r = 0, b = 0, l = 0, "cm")
  ) 
)
```

```{r, include = FALSE}
# Export plot as a jpeg file
ggsave(
  filename = "output/fig 3 annual abundance.jpeg",
  plot     = abun_year, 
  height   = 80, 
  width    = 200, 
  units    = "mm", 
  dpi      = 600
)
```

## Abundance and temperature (Figure 4)

Here we overlay a line graph of FF abundance since 2012 with maximum and minimum temperature (Fig 5) from the [Australian Bureau of Meteorology](http://www.bom.gov.au/climate/data/index.shtml?bookmark = 136) (navigate to `ACT` > `Past weather` > `Data and graphs` > `Text search` > `Temperature` > `Maximum temperature` or `Minimum temperature` > weather station `70351`). 

### Data preparation

```{r}
# Read in weather data
weather <- read_excel(raw_data, sheet = "weather") %>%
  clean_names() %>% 
  filter(year > 2011) %>%
  rename(
    max = "max_temperature_c", 
    min = "min_temperature_c"
  ) %>%
  
  # Format date
  mutate(date = paste(year, month, day, sep = "/")) %>%
  mutate(date = as.Date(date, format = c("%Y/%m/%d")))

# Pivot to long format
weather_long <- pivot_longer(
  weather,
  5:6, 
  names_to  = "type", 
  values_to = "temperature"
)
```

### Plot

```{r}
# Draw ribbon and line plot
(survey_weather_plot <- ggplot() +
   
  geom_ribbon(
    data = weather,
    aes(
      x    = as.Date(date),
      ymin = min * FF_SCALE,
      ymax = max * FF_SCALE
    ),
    alpha = 0.7,
    fill  = "#CACBE8"
  ) +
   
  geom_line(
    data = dat_all,
    aes(
      x = as.Date(date),
      y = bats_total
    ),
    colour   = "#1F2148",
    linewidth = 0.6
  ) +
   
  scale_x_date(
    name        = "Year",
    date_labels = "%Y",
    date_breaks = "2 years"
  ) +
   
  scale_y_continuous(
    name     = "Flying-fox abundance",
    breaks   = seq(0, 10000, 2000), 
    sec.axis = sec_axis(
      ~ . / FF_SCALE,
      name = "Daily temperature (°C)"
    )
  ) +
  
  # Apply visual theme
  theme_minimal() +
  theme(
    axis.line        = element_line(colour = "darkgrey"),
    axis.text.x      = element_text(vjust  = 1),
    axis.text.y      = element_text(vjust  = 0.5),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(vjust  = 1),
    legend.position  = "bottom",
    legend.text      = element_text(size   = 8),
    legend.title     = element_blank(),
    panel.background = element_rect(colour = NA, fill = "white"),
    panel.border     = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin      = margin(t = -0.3, r = 0, b = 0, l = 0, "cm")
  )
)
```

```{r, include = FALSE}
# Export plot as a jpeg file
ggsave(
  filename = "output/fig 4 abundance by temperature.jpeg",
  plot     = survey_weather_plot, 
  height   = 80, 
  width    = 200, 
  units    = "mm", 
  dpi      = 600
)
```

## Density maps (Figure 5)

Here we map the density of FFs in each tree across the camp (Fig 6). `surveys_wide` is from the `surveys` table in the database, and `trees` is from the `trees` table. 

Because we didn't do any surveys in June and July, but want them to appear in our figure, *you will need to add dummy surveys for these months* into the `surveys` dataframe. You can do this by copying the row for Aug 22, 2020 (where all the trees have 0 bats) several times for each missing month per year, and changing the date to one that occurs in each missing month (e.g., Jun 1, 2020 for June). 

### Data preparation

  1. Convert tree columns from wide format (one per column) to long format (tree identities listed in one column).

```{r}
# Read in data
surveys_wide <- read_excel(raw_data, sheet = "surveys") %>%
  subset(SurveyID != "N/A")

# Pivot the table to wide format
surveys <- melt(
  surveys_wide, 
  measure.vars  = c(30:203), 
  variable.name = "tree", 
  value.name    = "bats"
)
```

  2. Merge surveys with coordinates for each tree
  3. Create `month` and `year` variables

```{r, results = 'hide', warning = FALSE, message = FALSE}
# Read in the tree-by-tree data
trees <- read_excel(raw_data, sheet = "trees") %>%
  clean_names() %>%
  mutate(tree = tree_id)

# Join survey and tree data
surveys_trees <- left_join(surveys, trees, by = "tree") %>%
  clean_names() %>%
  subset(bats != 999) %>%
  mutate(
    year  = format(date, "%Y"),
    month = format(date, "%b"), 
    dates = format(date, "%d %b %y")
  )
```

 4. Get base layer from Google Maps

```{r, results = 'hide', warning = FALSE, message = FALSE}
# Fetch Google basemap
map <- get_map(
  location = c(lon = 149.131749, lat = -35.289492), 
  zoom     = 19, 
  source   = "google", 
  maptype  = "roadmap", 
  crop     = FALSE
)
```

  5. Order months
  6. Categorise number of bats
  7. Subset the data for 2019–23

```{r}
# Prepare the survey and tree data
data <- surveys_trees %>%
  
  # Format date column
  mutate(date = as.Date(date, format = c("%d/%m/%Y"))) %>%
  
  # Select only dates during the current reporting period
  subset(date > START & date < END) %>%
  
  mutate(
    month = factor(month, levels = c(
      "Dec", "Jan", "Feb", 
      "Mar", "Apr", "May", 
      "Jun", "Jul", "Aug", 
      "Sep", "Oct", "Nov"
    )), 
    
    foxes    = cut(
      as.numeric(bats), 
      breaks = c(-Inf, 50, 100, 200, 400, Inf), 
      labels = c(
        "0-50", 
        "50-100", 
        "100-200", 
        "200-400", 
        ">400"
      ))) %>%
  
  mutate(foxes = factor(foxes, levels = c(
    ">400", 
    "100-200", 
    "200-400", 
    "50-100", 
    "0-50"
  )))
```

### Plot

  6. Map the number of `Bats` recorded per `Month`

```{r}
# Plot density map
density_map <- ggmap(map) + 
  geom_point(
    data, 
    mapping = aes(x = longitude, y = latitude), 
    alpha   = 0.2, 
    col     = "darkgreen", 
    shape   = 8, 
    size    = 1
  ) +
  
  geom_point(
    data, 
    size    = 1, 
    mapping = aes(
      alpha = as.numeric(bats), 
      col   = as.numeric(bats), 
      x     = longitude, 
      y     = latitude
    )) +
  
  facet_wrap(~month, ncol = 3) +
  
  guides(
    alpha     = guide_legend(
      title   = "Number of \nflying-foxes\nrecorded", 
      reverse = TRUE), 
    col       = guide_legend(
      title   = "Number of \nflying-foxes\nrecorded", 
      reverse = TRUE)
    ) +
  
  labs(x = "", y = "") + 
  
  scale_colour_gradient2(
    low      = 0, 
    midpoint = 10, 
    mid      = "yellow", 
    high     = "red"
  ) + 
    
  # Apply visual theme
  theme(
    axis.line        = element_line(colour = "darkgrey"), 
    axis.text.x      = element_blank(), 
    axis.text.y      = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.title.x     = element_text(vjust = -2), 
    axis.title.y     = element_text(vjust = 5, margin = margin(r = 5)), 
    legend.key       = element_blank(), 
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 10), 
    panel.grid       = element_blank(),
    panel.background = element_rect(fill = "white", colour = NA), 
    plot.margin      = margin(t = -10, r = 4, b = 5, l = 5), 
    strip.background = element_rect(fill   = "#1F2148"),
    strip.text       = element_text(colour = "white")
  )

# Display map
# print(density_map)
```

```{r, include = FALSE, eval = FALSE}
# Save the plot as a jpeg file
ggsave(
  filename = "output/fig 6 tree density maps.jpeg", 
  plot     = density_map, 
  height   = 3500, 
  width    = 4000, 
  units    = "px", 
  dpi      = 800
)
```

## Camp occupation (Figure 6)

Here we plot the dates that flying-foxes occupied the camp each year. Because the flying-foxes never vacated the camp in 2014, we entered back-to-back dates to enable plotting. 

### Data preparation

```{r}
# Read in dates of camp occupation
occ <- read_excel(raw_data, sheet = "occupation") %>%
  clean_names() %>%
  
  # Exclude years with missing dates
  filter(!is.na(arrival) & !is.na(departure)) %>% 
  mutate(across(c(
    peak_date, 
    departure, 
    arrival, 
    year_start, 
    year_end), 
  as.Date),
         
  # Extract month and day and apply a single year
  arr_date  = as.Date(paste("2012", format(arrival,   "%m-%d"), sep = "-")),
  dep_date  = as.Date(paste("2012", format(departure, "%m-%d"), sep = "-")),
  peak_date = as.Date(paste("2012", format(peak_date, "%m-%d"), sep = "-")),
  
  # Treat year as a factor
  year = factor(year),
  
  # Make another year factor for a secondary y-axis
  year_sec = as.numeric(as.character(year)))
```

### Plot

```{r}
# Generate plot
(occ_plot <- ggplot(occ) +
  geom_segment(
    aes(
      color   = year, 
      y       = year_sec, 
      yend    = year_sec, 
      x       = as.Date(format(year_start, "2012-%m-%d")),
      xend    = as.Date(format(  dep_date, "2012-%m-%d"))
    ), 
    linewidth = 2
  ) +
  
  geom_segment(
    aes(
    color     = year, 
    x         = as.Date(format(arr_date, "2012-%m-%d")),
    xend      = as.Date(format(year_end, "2012-%m-%d")), 
    y         = year_sec, yend = year_sec
    ), 
    linewidth = 2
  ) +
  
  geom_point(
    aes(
      fill = year, 
      x = as.Date(format(peak_date, "2012-%m-%d")), 
      y = year_sec 
    ), 
    alpha  = 0.6, 
    color  = "white", 
    shape  = 21, 
    size   = 5, 
    stroke = 1.2
  ) +
 
   # Add departure date labels after the bar
  geom_text(
    aes(
      label = format(dep_date, "%d %b"), 
      x     = as.Date(format(dep_date, "2012-%m-%d")) + 5, 
      y     = year_sec
    ),
    color   = "grey50", 
    hjust   = 0, 
    size    = 2.5
  ) +
 
   # Add arrival date labels before the bar
  geom_text(
    aes(
      label = format(arr_date, "%d %b"), 
      x     = as.Date(format(arr_date, "2012-%m-%d")) - 5, 
      y     = year_sec
    ),
    color   = "grey50", 
    hjust   = 1, 
    size    = 2.5
  ) +
 
   scale_x_date(
     date_labels = "%b", 
     date_breaks = "1 month",
     limits      = c(
       as.Date("2012-01-01"), 
       as.Date("2012-12-31")
  )) +
  
  labs(x = "Month") + 
  scale_color_viridis_d() + 
  scale_fill_viridis_d() + 
  scale_y_continuous(
    name     = "Year", 
    breaks   = occ$year_sec, 
    labels   = occ$year, 
    sec.axis = sec_axis(
      ~., 
      breaks = occ$year_sec, 
      labels = occ$peak_abundance, 
      name   = "Peak flying-fox abundance"
  )) +
  
  # Apply visual theme
  theme_minimal() +
  theme(
    axis.line          = element_line(colour = "darkgrey"), 
    axis.text.x        = element_text(hjust = 1), 
    axis.text.y        = element_text(hjust = 1), 
    axis.title.x       = element_blank(), 
    axis.title.y       = element_text(margin = margin(r = 10)), 
    axis.title.y.right = element_text(margin = margin(l = 10)), 
    legend.position    = "none", 
    legend.text        = element_text(size = 8),
    legend.title       = element_blank(), 
    panel.background   = element_rect(fill = "white", colour = NA), 
    panel.grid         = element_blank(),
    plot.margin        = margin(t = 0, r = 0, b = 0, l = 0)
  )
)
```

```{r, include = FALSE}
# Save the plot as a jpeg file
ggsave(
  filename = "output/fig 6 occupation.jpeg",
  plot     = occ_plot, 
  height   = 3500, 
  width    = 6000, 
  units    = "px", 
  dpi      = 800
)
```

## Categorical variables (Figure 7)

### Data preparation

```{r}
# Prepare levels counts for categorical variables
dat_cat <- dat_cur %>%
  
  # Remove estimates
  filter(survey == "Count") %>%
  
  # Rename categorical variables
  rename(
    heat  = heat_stress, 
    humid = humidity, 
    pups  = mothers_and_pups
  ) %>%
  
  # Format month
  mutate(month = floor_date(date, "month")) %>%
  
  # Transform to long format
  pivot_longer(
    cols      = all_of(cat_vars),
    names_to  = "var",
    values_to = "level"
  ) %>%
  
  # Remove NAs
  filter(!is.na(level), level != "") %>%
  
  # Sum levels per variable
  count(month, var, level)
```

### Plot

```{r}
# Barplot categorical variables
cat_plots <- dat_cat %>%
  
  # Group by categorical variable
  split(.$var) %>%
    purrr::map(~ {
      v <- unique(.x$var)

    ggplot(.x, aes(x = month, y = n, fill = level)) +
      
      # Draw bar plot
      geom_col() +
      
      # Define labels
      labs(
        fill  = "",
        title = cat_labs[[v]], 
        x     = "Month",
        y     = "Frequency"
      ) +
      
      # Define fill colours
      scale_fill_manual(
        values = c(
          "#2C2F69", 
          "#8E92DE", 
          "#A1CC76", 
          "#C4A487" 
      )) +
      
      # Define date axis
      scale_x_date(
        date_labels = "%b",
        date_breaks = "1 month",
        expand      = c(0, 0)
      ) +
      
      # Apply visual theme
      theme_minimal() +
      theme(
        axis.line        = element_line(colour = "darkgrey"),
        axis.text.x      = element_text(angle  = 90, vjust = 1),
        axis.text.y      = element_text(vjust  = 0.5),
        axis.title.x     = element_blank(),
        axis.title.y     = element_text(vjust  = 3, margin = margin(r = 1)),
        legend.position  = "bottom",
        legend.text      = element_text(size   = 9),
        legend.title     = element_blank(),
        panel.background = element_rect(colour = NA, fill = "white"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border     = element_blank(),
        plot.margin      = margin(t = 0.4, r = 0.5, b = 0.5, l = 0.5, "cm"), 
        plot.title       = element_text(face = "bold", hjust = 0.5, size = 12)
      )
  }
)

# Display plots
print(cat_plots$pups)
print(cat_plots$heat)
print(cat_plots$bat_noise)
print(cat_plots$other_noise)
print(cat_plots$rubbish)
print(cat_plots$humans)
print(cat_plots$cloud)
print(cat_plots$humid)
print(cat_plots$wind)
```

```{r, include = FALSE}
# Reorder plots
cat_plots <- cat_plots[c(
  "pups",
  "heat",
  "bat_noise",
  "other_noise",
  "humans",
  "rubbish",
  "cloud",
  "humid",
  "wind"
)]

# Combine plots
(cat_plots_comb <- ggarrange(
  plotlist      = cat_plots,
  ncol          = 3,
  nrow          = ceiling(length(cat_plots) / 3), 
  common.legend = FALSE
  )
)

# Export plot as a jpeg file
ggsave(
  filename = "output/fig 5 categories.jpeg",
  plot     = cat_plots_comb, 
  height   = 200, 
  width    = 350, 
  units    = "mm", 
  dpi      = 600
)
```

# **Session information**

```{r, eval = TRUE}
# Display R version, OS, and loaded packages
sessionInfo()
```