---
title: "ABS Commonwealth Park flying-fox population analyses and mapping"
author: "Belinda Wilson and Shoshana Rapley"
date: "December 22, 2024"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
    theme: cerulean
    highlight: pygments
editor_options:
  
  chunk_output_type: console
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'tutorial.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE)
```

# Background

In the Australian Capital Territory (ACT), [Commonwealth Park](https://www.nca.gov.au/attractions/commonwealth-park#) has been occupied by flying-foxes (FFs) seasonally every year since 2003, and the [Australasian Bat Society Inc](https://www.ausbats.org.au/) (ABS) has been surveying the camp since then. In 2012, the ABS entered into an agreement with the [National Capital Authority](https://www.nca.gov.au/environment/national-land/conservation-land-management/commonwealth-park-grey-headed-flying-fox) (NCA) to monitor this camp. Since 2013, the ABS has also been providing quarterly census survey data to the ACT and Commonwealth Governments as part of the [National Flying-fox Monitoring Program](https://www.dcceew.gov.au/environment/biodiversity/threatened/species/flying-fox-monitoring#:~:text=The%20National%20Flying%2Dfox%20Monitoring%20Program%20(NFFMP)%20is%20designed,flying%2Dfoxes%20in%20eastern%20Australia.). The survey methodology is consistent with the [CSIRO Commonwealth Flying-fox census method](http://www.environment.gov.au/biodiversity/threatened/species/pubs/310112-monitoring-methodology.pdf).

Here we present the abundance and density results generated as part of the ABS' annual report to the NCA. Note that the reporting period for each year extends Dec 1–Nov 30, and the species detected were either **grey-headed flying-fox** (GHFF) or **little red flying-fox** (LRFF).

# Data preparation

We enter the results of FF surveys in a [Microsoft Access](https://www.microsoft.com/en-au/microsoft-365/access) database. To put this into a form that we can use in this R markdown, copy and paste the `Survey` and `Tree` tables from the database into the `input/data.xlsx` sheets of the same name. 

Then, copy and paste the rows from the last reporting period (Dec 1–Nov 30) in the `Schedule` sheet of the `ABS flying-fox monitoring` Google spreadsheet (requires specific access) to the bottom of the `data.xlsx`'s `monitoring` sheet.

# Setup

We used the [pacman Package Management Tool](https://cran.r-project.org/web/packages/pacman/index.html) to install and load required packages in a condensed and efficient way.

```{r, eval=FALSE}
#install.packages("pacman")
```

```{r, results='hide', warning=FALSE, message=FALSE}
# Install and load required packages
pacman::p_load(plyr, data.table, dplyr, ggmap, ggpubr, ggplot2, janitor, lubridate, naniar, readxl, renv, reshape2, tidyr, tidyverse)
```

Here we set the constants for this markdown.

```{r}
# Assign raw data filename to an object
raw_data <- "input/data.xlsx"

# Set dates for this year's reporting period
START <- "2023-12-01"
END <- "2024-11-30"
CURRENT_YEAR <- 2024
```

# Species and demographics

First, we generate a **(a)** raw line graph and a **(b)** smoothed line graph by month, with two y-axes (Fig 3a):

  1. Number of GHFFs over time (primary axis ranging 0–10,000)
  2. Number of LRFFs over time (primary axis ranging 0–10,000)
      - Currently not required, since no LRFFs were recorded this year.
  3. Number of trees occupied (secondary axis ranging 0–160)

```{r, results='hide', warning=FALSE, message=FALSE}
# Prepare data
data <- read_excel(raw_data, sheet="monitoring") %>%
  clean_names() %>%
  filter(include=="Yes", location=="Commonwealth Park")%>%
  mutate(lrff=as.character(lrff)) %>%
  pivot_longer(cols=9:10, names_to="species", values_to="count") %>%
  #filter(count!=0) %>% 
  mutate(count=if_else(count=="800-1000", "900", count)) %>%
  mutate(count=if_else(count=="200-250", "225", count)) %>% 
  #replaces ranges with mean
  mutate(count=as.numeric(count),
         trees_occupied=as.numeric(trees_occupied),
         date=as.Date(date, format=c("%d/%m/%Y"))) %>%
  mutate(year=year(date))

# Subset to current reporting period
data_current <- data %>%
  filter(date > START & date < END) %>%
  # If no LRFFs were recorded this year, we removed these
  filter(species != "lrff")
```

**Raw line graph**

```{r}
# Line plot raw abundance
abun_raw <- ggplot(data_current) +
  geom_line(mapping=aes(date, count, col=species), linewidth=0.6) +
  geom_line(mapping=aes(date, trees_occupied*50, 
                col='chartreuse4'), linewidth=0.6) +
  scale_color_manual(name='', 
                     values=c("#CACBE8","#1F2148","#7B7FC7"), 
                     labels=c("Trees occupied", 
                              "Grey-headed flying-fox")) +
  scale_x_date(date_breaks="1 month", date_labels="%b") +
  scale_y_continuous(breaks=seq(0, 10000, 2000), 
                     sec.axis=sec_axis(~ ./50, 
                      name="Trees occupied")) +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),
        legend.text=element_text(size=8),
        legend.position="bottom", 
        legend.title=element_blank(),
        axis.text.y=element_text(angle=0, 
                    vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=0, 
                    vjust=1, hjust=0, colour="black"),
        axis.line=element_line(colour="lightgrey"), 
        # Add space to the right for combined plot
        plot.margin=margin(t=0.2, r=0.5, b=0, l=0.1, "cm")) +
  labs(x="Month", y="Flying-foxes recorded")

# Display the plot
print(abun_raw)
```

**Smoothed line graph**

```{r}
# Smooth plot abundance
abun_smooth <- ggplot(data_current) +
  geom_smooth(mapping=aes(date, count, col=species, fill=species), 
              linewidth=0.6) +
  geom_smooth(mapping=aes(date, trees_occupied*50, 
                col='chartreuse4', fill='chartreuse4'), linewidth=0.6) +
  scale_color_manual(name='', values=c("#CACBE8","#1F2148","#7B7FC7"), 
                     labels=c("Trees occupied", 
                              "Grey-headed flying-fox")) +
  scale_fill_manual(name='', values=c("#CACBE8","#1F2148","#7B7FC7"), 
                     labels=c("Trees occupied", 
                              "Grey-headed flying-fox")) +
  scale_x_date(date_breaks="1 month", date_labels="%b") +
  scale_y_continuous(breaks=seq(0, 10000, 2000), 
                     sec.axis=sec_axis(~ ./50, 
                      name="Trees occupied")) +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),
        legend.text=element_text(size=8),
        legend.position="bottom", 
        legend.title=element_blank(),
        axis.text.y=element_text(angle=0, 
                    vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=0, 
                    vjust=1, hjust=0, colour="black"),
        axis.line=element_line(colour="lightgrey"), 
        # Add space to the left for combined plot
        plot.margin=margin(t=0.2, r=0.1, b=0, l=0.5, "cm")) +
  labs(x="Month", y="Flying-foxes recorded")

# Display the plot
print(abun_smooth)
```

```{r}
# Create a light grey dividing line as an empty plot
divider <- ggplot() +
  # Add the dividing line
  geom_vline(xintercept=0, color="black", size=0.5) +
  # Extend the line infinitely in the y-direction
  ylim(-Inf, Inf) +
  theme_void()

# Combine the plots
abun_current <- ggarrange(abun_raw, divider, abun_smooth, 
                          ncol=3, nrow=1, labels=c("a", "", "b"), 
                          label.x=0.8, label.y=0.98, 
                          widths=c(1, 0.02, 1), align="h")

# Display the combined plot
print(abun_current)
```

```{r, include=FALSE}
# Save the plot as a jpeg file
ggsave(plot=abun_current, 
       filename="output/fig 3a (species abundance).jpeg",
       height=100, width=250, units="mm", dpi=800)
```

# Mean abundance

Bar graph with month on the x-axis and FF abundance on the y-axis, showing the following as separate bars with standard error bars (Fig 3b): 

  1. Mean (±SE) FFs recorded (primary axis ranging 0–6,000)
  2. Mean (±SE) trees occupied (secondary axis ranging 0–140)
  3. Mean (±SE) ratio of FFs-to-trees (secondary axis ranging 0–140)

For the purposes of plotting the mean number of trees occupied and tree-by-bat ratio on a secondary axis, we multiplied these values by 40, then reversed this before plotting.

```{r}
# Read in the data
data_current <- read_excel(raw_data, sheet="monitoring") %>% 
  clean_names() %>% 
  filter(include=="Yes", location=="Commonwealth Park") %>% 
  mutate(year=year(date)) %>% 
  subset(date > START & date < END) %>% 
  mutate(month=lubridate::month(date, label=TRUE, abbr=TRUE), 
         ghff=as.numeric(ghff), 
         lrff=as.numeric(lrff), 
         bats_total=ghff + lrff, 
         ratio=as.numeric(bats_total/trees_occupied)) %>% 
  mutate(ratio=ifelse(ratio=="NaN", 0, ratio)) 

# Calculate mean statistics
data_current_means <- data_current %>% 
  group_by(month) %>% 
  summarise(count_mean=mean(bats_total),
            ratio_mean=40*mean(ratio),
            trees_mean=40*mean(trees_occupied)) %>%
  pivot_longer(2:4, names_to="type", values_to="mean")

# Calculate standard error (SE) statistics
data_current_se <- data_current %>% 
  group_by(month) %>% 
  summarise(count_se=sd(bats_total)/sqrt(length(bats_total)),
            ratio_se=40*sd(ratio)/sqrt(length(ratio)),
            trees_se=40*sd(trees_occupied)/sqrt(length(trees_occupied))) %>%
  pivot_longer(2:4, names_to="type", values_to="se")

# Combine means and SE
data_current_stats <- cbind(data_current_means, data_current_se)
colnames(data_current_stats) <- make.unique(names(data_current_stats))
```

So each month plots regardless of whether FFs were detected, manually generate rows for these months.

```{r}
# Manually create rows for months where bats were not detected
jul <- data.frame("Jul", "count_mean", NA, "Jul", "count_se", NA)
names(jul) <- c("month", "type", "mean", "month.1", "type.1", "se")

aug <- data.frame("Aug", "count_mean", NA, "Aug", "count_se", NA)
names(aug) <- c("month", "type", "mean", "month.1", "type.1", "se")

sep <- data.frame("Sep", "count_mean", NA, "Sep", "count_se", NA)
names(sep) <- c("month", "type", "mean", "month.1", "type.1", "se")

# Combine rows with df
data_current_stats <- rbind(data_current_stats, 
                            jul, aug, sep) %>%
  mutate(type=factor(type,levels=c("count_mean", "trees_mean", 
                                   "ratio_mean", "count_se", 
                                   "trees_se", "ratio_se")),
         month=factor(month, levels=c("Dec", "Jan", "Feb", "Mar", 
                                      "Apr", "May", "Jun", "Jul", 
                                      "Aug", "Sep", "Oct", "Nov")))

# Reverse ×40 multiplier
data_current_stats$true <- data_current_stats$mean/40 
```

```{r}
# Plot
mean_se_plot <- ggplot(data_current_stats, 
                       aes(x=month, y=mean, fill=type)) +
  geom_bar(stat="identity", position="dodge") + 
  geom_errorbar(mapping=aes(x=month, ymin=mean-se, 
                            ymax=mean+se, col=type), 
                position="dodge") + 
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),
        legend.text=element_text(size=8),
        legend.position="bottom", 
        legend.title=element_blank(),
        axis.text.y=element_text(angle=0, 
                    vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=0, 
                    vjust=1, hjust=0, colour="black"),
        plot.margin=margin(0.5, 0, 0, 0, "cm")) +
  scale_fill_manual(name='', values=c("#1F2148","#CACBE8","#7B7FC7"),
                    labels=c("Mean flying-fox abundance", 
                             "Mean trees occupied", 
                             "Mean ratio of flying-fox to trees")) +
  scale_y_continuous(sec.axis=sec_axis(~ ./40, 
                    name="Mean trees occupied and \nmean ratio of flying-fox to trees")) +
  scale_colour_manual(name='', guide="none",
                    values=c("#CACBE8","#7B7FC7","#1F2148")) +
  labs(x="Month", y="Mean flying-fox abundance")

# Display the plot
print(mean_se_plot)
```

```{r, include=FALSE}
# Save the plot as a jpeg file
jpeg(file="output/fig 3b (mean abundance).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(mean_se_plot)
dev.off()
```

# Peak numbers

Stacked bar graph for 2012-2023 with year on the x-axis and FF abundance on the y-axis (Fig 4), showing:

  1. Peak (maximum) FFs recorded for each year
  2. Mean FFs recorded for each year

```{r}
# Prepare the data
data <- read_excel(raw_data, sheet="monitoring") %>%
  clean_names() %>%
  drop_na(ghff)%>%
  filter(include=="Yes", location=="Commonwealth Park") %>%
  mutate(ghff=if_else(ghff=="800-1000", "900", ghff)) %>%
  # Replace ranges with mean
  mutate(ghff=if_else(ghff=="200-250", "225", ghff)) %>% 
  mutate(ghff=as.numeric(ghff),
         lrff=as.numeric(lrff),
         year=year(date)) %>% 
  mutate(bats_total=ghff + lrff) %>%
  # Remove years prior to year 2000
  filter(year > 2000)

# Subset to peak abundances for each year
peak <- data %>% 
  group_by(year) %>%
  summarise(bats_mean=mean(bats_total), 
            bats_max=max(bats_total)) %>%
  mutate(bats_diff=bats_max-bats_mean) %>%
  pivot_longer(2:4, names_to="type", values_to="stat") %>%
  filter(type!="bats_max")
```

```{r}
# Bar plot
abun_years <- ggplot(peak, (aes(year, stat, fill=type))) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),
        legend.text=element_text(size=8),
        legend.position="bottom", 
        legend.title=element_blank(),
        axis.text.y=element_text(angle=0, 
                    vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=0, 
                    vjust=1, hjust=0, colour="black"),
        plot.margin=margin(0.5, 0, 0, 0, "cm")) +
  scale_fill_manual(name="", values=c("#1F2148","#CACBE8"),
                    labels=c("Peak", "Mean")) +
  scale_y_continuous(breaks=seq(0, 10000, 2000)) +
  scale_x_continuous(breaks=seq(2011, CURRENT_YEAR, 1)) +
  labs(x="Year", y="Flying-fox abundance") +
  theme(plot.title=element_text(size=15, face="bold", 
                                margin=margin(10, 0, 10, 0)))

# Display the plot
print(abun_years)
```

```{r, include=FALSE}
# Save plot as a jpeg
jpeg(file="output/fig 4 (abundance across years).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(abun_years)
dev.off()
```

# Survey effort and weather

Here we generate a line graph of FFs recorded across all years (2012–current), as well as temperature in the background (Fig 5). 

We acquired weather data from the [Australian Bureau of Meteorology](http://www.bom.gov.au/climate/data/index.shtml?bookmark=136) by navigating to ACT > Past weather > Data and graphs > Text search > Maximum temperature or Minimum temperature in weather station 70351. 

```{r}
# Read in weather data
weather <- read_excel(raw_data, sheet="weather") %>%
  clean_names() %>% 
  filter(year > 2011) %>%
  rename(max="maximum_temperature_degree_c", 
         min="minimum_temperature_degree_c") %>%
  mutate(date=paste(year, month, day, sep="/")) %>%
  mutate(date=as.Date(date, format=c("%Y/%m/%d")))

# Pivot to long format
weather_long <- pivot_longer(weather, 5:6, 
                         names_to="type", 
                         values_to="temperature")

# Plot
survey_weather_plot <- ggplot() +
  geom_ribbon(data=weather, aes(as.Date(date), 
                                ymin=min, ymax=max), 
              fill="#CACBE8", alpha=0.5) +
  geom_line(data=data, aes(as.Date(date), bats_total/200), 
            size=0.6, col="#1F2148") +
  theme_minimal() +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank(), 
        panel.border=element_blank(),
        legend.text=element_text(size=8),
        legend.position="bottom", 
        legend.title=element_blank(),
        axis.text.y=element_text(angle=0, 
                    vjust=0.5, colour="black"),
        axis.text.x=element_text(angle=0, 
                    vjust=1, hjust=0, colour="black"),
        plot.margin=margin(0.5, 0, 0, 0, "cm")) +
  scale_x_date(name="Year", date_labels="%Y", 
               date_breaks="2 years") +
  scale_y_continuous(name="Daily temperature (°C)", 
                     sec.axis=sec_axis(~ .*200, 
                                       breaks=seq(0, 10000, 2000),
                                       name="Flying-fox abundance"))

# Display the plot
print(survey_weather_plot)
```

```{r, include=FALSE}
# Save plot as a jpeg
jpeg(file="output/fig 5 (surveys and weather).jpeg", 
     width=6000, height=2500, units="px", res=800)
print(survey_weather_plot)
dev.off()
```

# Density map

Here we map the density of FFs in each tree across the camp (Fig 6). `surveys_wide` is from the `surveys` table in the database, and `trees` is from the `trees` table. 

Because we didn't do any surveys in June and July, but want them to appear in our figure, *you will need to add dummy surveys for these months* into the `surveys` dataframe. You can do this by copying the row for Aug 22, 2020 (where all the trees have 0 bats) several times for each missing month per year, and changing the date to one that occurs in each missing month (e.g., Jun 1, 2020 for June). 

  1. Convert tree columns from wide format (one per column) to long format (tree identities listed in one column).

```{r}
# Read in data
surveys_wide <- read_excel("input/data.xlsx", sheet="surveys") %>%
  subset(SurveyID!="N/A")

# Pivot the table to wide format
surveys <- melt(surveys_wide, measure.vars=c(30:203), 
                variable.name="tree", value.name="bats")
```

  2. Merge surveys with coordinates for each tree
  3. Create `month` and `year` variables

```{r, results='hide', warning=FALSE, message=FALSE}
# Read in the tree-by-tree data
trees <- read_excel(raw_data, sheet="trees") %>%
  clean_names() %>%
  mutate(tree = tree_id)

# Join survey and tree data
surveys_trees <- left_join(surveys, trees, by="tree") %>%
  clean_names() %>%
  subset(bats!=999) %>%
  mutate(year=format(date, "%Y"),
         month=format(date, "%b"), 
         dates=format(date, "%d %b %y"))
```

 4. Get base layer from Google Maps

You'll need to sign up for a Google API for this step.

```{r, results='hide', warning=FALSE, message=FALSE}
# Use a Google API to fetch a base map
#ggmap::register_google(key="[enter your key here]", write=TRUE)
map <- get_map(location=c(lon=149.131749, lat=-35.289492), 
               zoom=19, source="google", 
               maptype="roadmap", crop=FALSE)
```

  5. Order months
  6. Categorise number of bats
  7. Subset the data for 2019–23

```{r}
# Prepare the survey and tree data
data <- surveys_trees %>%
  mutate(date=as.Date(date, format=c("%d/%m/%Y"))) %>%
  subset(date > START & date < END) %>%
  mutate(month=factor(month, levels=c("Dec", "Jan", "Feb", 
                                      "Mar", "Apr", "May", 
                                      "Jun", "Jul", "Aug", 
                                      "Sep", "Oct", "Nov")),
         foxes=cut(as.numeric(bats), 
                   breaks=c(-Inf, 50, 100, 200, 400, Inf), 
                   labels=c("0-50", "50-100", "100-200", 
                            "200-400", ">400"))) %>%
  mutate(foxes=factor(foxes, levels=c(">400", "100-200", 
                                      "200-400", "50-100", 
                                      "0-50")))
```

  6. Map the number of `Bats` recorded per `Month`

```{r}
# Plot density map
density_map <- ggmap(map) + 
  geom_point(data, mapping=aes(x=longitude, y=latitude), 
             shape=8, col="darkgreen", size=1, alpha=0.2) +
  geom_point(data, size=1, 
             mapping=aes(x=longitude, y=latitude, 
                         col=as.numeric(bats), alpha=as.numeric(bats))) +
  scale_colour_gradient2(low=0, midpoint=10, 
                         mid="yellow", high="red") +
  facet_wrap(~month, ncol=3) +
  theme(panel.grid.major=element_blank(),
        panel.background=element_rect(fill="white"),
        plot.margin=unit(c(0, 0, 0, 0), "cm"),
        axis.text.x=element_blank(), 
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        legend.key=element_blank(),
        strip.background=element_rect(fill="#1F2148"),
        strip.text=element_text(colour="white")) +
  guides(col=guide_legend(title="Number of \nflying-foxes\nrecorded",
                          reverse=TRUE), 
         alpha=guide_legend(title="Number of \nflying-foxes\nrecorded",
                            reverse=TRUE)) +
  xlab("") + ylab("")

# Display the plot
print(density_map)
```

```{r, include=FALSE}
# Save the plot as a jpeg file
ggsave(plot=density_map, 
       filename="output/fig 6 (density maps).jpeg",
       width=4000, height=3500, units="px", dpi=800)
```

# Session information

```{r}
# Display version information for R, OS, and packages
sessionInfo()
```